---
title: "Class 9: Candy Mini-Project"
author: "Hailey Phan (PID: A19113445)"
format: pdf
toc: true
---

## Background 
Today we are taking a detor to analyze a fun dataset (that we have more intrinsic insight into) with the most useful analysis method we have learned thus far - Principal Component Analysis (PCA).

## Data Import
The data is all related to Halloween candy and is from the 538 website.

```{r}
candy_file <- "candy-data.txt"
candy <- read.csv(candy_file, row.names = 1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```


> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

## What is your favorite candy?

```{r}
candy["Twix", ]$winpercent
```

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Twix") |> 
  select(winpercent)
```
> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

My favorite candy is 100 grand. 

```{r}
candy["100 Grand", ]$winpercent
```


> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```


## Side-note: the `skimr::skim()` function
There is a useful `skim()` function in the **skimr** package that can help give you a quick overview of a given dataset. Let’s install this package and try it on our candy data.

```{r}
library("skimr")
skim(candy)
```

> Q6.  Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, winpercent is on a different scale because it ranges roughly from 20–85, while most columns are 0/1.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

In chocolate, 1 means the candy contains chocolate and 0 means it does not.


## Exploratory analysis
A good place to start any exploratory analysis is with a histogram. You can do this most easily with the base R function `hist()`. Alternatively, you can use `ggplot()` with `geom_hist()`. Either works well in this case and (as always) its your choice.

> Q8. Plot a histogram of winpercent values using both base R and ggplot2.

```{r}
hist(candy$winpercent,
     main = "Histogram of Candy Win Percent",
     xlab = "Win Percent",
     col = "lightblue")
```

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black") +
  labs(title = "Histogram of Candy Win Percent",
       x = "Win Percent",
       y = "Count")
```

> Q9. Is the distribution of winpercent values symmetrical?

```{r}
# Summary stats to judge center
summary(candy$winpercent)

# Mean and median (helps determine symmetry)
mean(candy$winpercent)
median(candy$winpercent)
```

No, it is slightly skewed to the right. (Since mean > median).

> Q10. Is the center of the distribution above or below 50%?

Since the mean = 50.32, the center of the distribution is slightly above 50%.

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# Mean winpercent for chocolate candies
mean_chocolate <- mean(candy$winpercent[as.logical(candy$chocolate)])

# Mean winpercent for fruity candies
mean_fruity <- mean(candy$winpercent[as.logical(candy$fruity)])

mean_chocolate
mean_fruity
```

Chocolate candies are ranked higher on average than fruity candies.

> Q12. Is this difference statistically significant?

```{r}
t.test(
  candy$winpercent[as.logical(candy$chocolate)],
  candy$winpercent[as.logical(candy$fruity)]
)
```

Based on our t-test, which yielded a small p-value of 2.871e-08, the difference is highly significant. So, chocolate candies are significantly more popular than fruity candies. 


## Overall Candy Rankings
Let’s use the base R `order()` function together with `head()` to sort the whole dataset by a variable/column of interest. Or if you have been getting into the tidyverse and the dplyr package you can use the `arrange()` function together with `head()` to do the same thing and answer the following questions:

> Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent), ], n = 5)
```


> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent, decreasing = TRUE), ], n = 5)
```


> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r, fig.height=15, fig.width=10}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r, fig.height=15, fig.width=10}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col() +
  ylab("Candy") +
  xlab("Win Percent")
```

## Add some color

```{r}
my_cols = rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate4"
my_cols[as.logical(candy$bar)] = "deeppink4"
my_cols[as.logical(candy$fruity)] = "pink"
```


```{r, fig.height=15, fig.width=10}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill = my_cols) +
  ylab("")
```

> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets.

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst.

## Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of win vs price
ggplot(candy) +
  aes(pricepercent, winpercent, label = rownames(candy)) +
  geom_point(col = my_cols) + 
  geom_text_repel(col = my_cols, size = 3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese’s Miniatures.

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
candy[ord, c("pricepercent", "winpercent")][1:5, ]
```
The least popular among these is the Nik L Nip because it has the lowest winpercent. 

## Exploring the correlation structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

The two variables that are anti-correlated are chocolate and fruity. This means that candies that contain chocolate are very unlikely to be fruity, and vice versa.

> Q23. Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?

The variables that are expected to have the largest contributions to PC1 are chocolate, bar, winpercent, and pricepercent in the positive direction, and fruity and hard in the negative direction. This suggests that the main variation in the dataset separates chocolate-based candies from fruity candies.

## Principal Component Analysis (PCA)

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
plot(pca$x[, 1:2])
```

```{r}
plot(pca$x[,1:2], col = my_cols, pch = 16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
  aes(x = PC1, y = PC2,
      size = winpercent/100,
      text = rownames(my_data),
      label = rownames(my_data)) +
  geom_point(col = my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")

```

```{r}
#library(plotly)

#ggplotly(p)
```

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

The variables that are picked up strongly by PC1 in the positive direction include chocolate, bar, winpercent, and pricepercent. This makes sense because chocolate candies tend to be more popular and often more expensive. This relationship was previously highlighted in the correlation matrix, where chocolate showed a positive correlation with winpercent and pricepercent.

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

A winning candy is typically chocolate-based, often in the form of a candy bar, and may include ingredients such as peanuts, caramel, or wafers. The visualizations show that top-ranked candies are mostly chocolate, while the correlation analysis demonstrates that chocolate is positively associated with popularity. The PCA further supports this by separating chocolate and fruity candies along the first principal component, confirming that chocolate candies tend to be more popular overall.