---
title: "Class 8 Mini-Project: Unsupervised Learning Analysis of Human Breast Cancer Cells"
author: "Hailey Phan (PID: A19113445)"
format: pdf
toc: TRUE
---

## Background

The goal of this mini-project is for you to explore a complete analysis using the unsupervised learning techniques covered in class.

Today we will analyze a data from fine needle aspiration (FNA) of a breast mass.

## Data Import

The data is made available as a CSV file for downlad. We can read this in using `read.csv()`:

```{r}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)
```

```{r}
head(wisc.df, 4)
```
Make sure we remove or exclude the `diagnosis` column from the dataset that we use for further analysis - this is the expert diagnosis as either M or B:
```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
diagnosis <- as.factor(wisc.df$diagnosis)
```

## Exploratory data analysis

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```


> Q2. How many of the observations have a malignant diagnosis?

```{r}
table(diagnosis)
```

> Q3. How many variables/features in the data are suffixed with `_mean`?

We can use the `grep()` function to help us here:

```{r}
length(grep("_mean", colnames(wisc.data)))
```


```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

## Principal Component Analysis (PCA)

We need to scale our data before PCA with the `scale = TRUE` argument to `prcomp()`.


```{r}
# Perform PCA on wisc.data
wisc.pr <- prcomp(wisc.data, scale = TRUE)

# Look at summary of results
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

0.4427 = 44.27%

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 PCs

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7 PCs

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

What stands out is that the plot looks very crowded and messy. It is difficult to understand because there are too many observations and variable arrows all overlapping in the same space, so it is hard to clearly see patterns, clusters, or which variables are contributing most to the principal components.

```{r}
biplot(wisc.pr)
```

Let's see the "PC score plot"

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col = diagnosis) +
  geom_point()
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col = diagnosis) +
  geom_point()
```
PC1 still shows a clear separation between malignant and benign samples, while PC3 shows much less separation.

## Variance explained


A scree plot shows how much variance each PC captures. We typically look for an “elbow” — a point where adding more PCs gives diminishing returns. This can help us decide how many PCs to consider for further analysis.

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```


```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

```{r}
## ggplot based graph
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

> Q9. For the first principal component, what is the component of the loading vector (i.e. `wisc.pr$rotation[,1]`) for the feature `concave.points_mean`? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```
Yes, there are features with slightly larger absolute contributions to PC1.


## Hierarchical clustering

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)

data.dist <- dist(data.scaled)

wisc.hclust <- hclust(data.dist, method = "complete")
```

> Q10. Using the `plot()` and `abline()` functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h = 19, col = "red", lty = 2)
```
About 19.


```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
table(wisc.hclust.clusters, diagnosis)
```


> Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

```{r}
plot(hclust(data.dist, method = "single"))
plot(hclust(data.dist, method = "complete"))
plot(hclust(data.dist, method = "average"))
plot(hclust(data.dist, method = "ward.D2"))
```
`ward.D2` is the best choice here because it gives cleaner, more balanced clusters.

## Combining methods

```{r}
## Use the distance along the first 7 PCs for clustering i.e. wisc.pr$x[, 1:7]
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method = "ward.D2")
```

```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k = 2)
table(wisc.pr.hclust.clusters, diagnosis)
```

> Q13. How well does the newly created `hclust` model with two clusters separate out the two “M” and “B” diagnoses?

```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```
This model separates the malignant and benign diagnoses quite well, with most malignant samples in one cluster and most benign samples in the other.

> Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the `table()` function to compare the output of each model (`wisc.hclust.clusters` and `wisc.pr.hclust.clusters`) with the vector containing the actual diagnoses.

```{r}
table(wisc.hclust.clusters, diagnosis)
```
The earlier clustering without PCA is less clean and shows more mixing between benign and malignant samples.

## Prediction

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
grps <- cutree(wisc.pr.hclust, k = 2)
npc
```
```{r}
plot(wisc.pr$x[,1:2], col = grps)
points(npc[,1], npc[,2], col = "blue", pch = 16, cex = 3)
text(npc[,1], npc[,2], c(1,2), col = "white")
```

> Q16. Which of these new patients should we prioritize for follow up based on your results?

Patient 1 should be prioritized for follow-up because it falls closer to the malignant cluster pattern in PCA space.











