---
title: "HW Class12 Pt.2 (Population analysis) [Q13 Q14 BoxPlot]"
author: "Hailey Phan (PID: A19113445)"
format: pdf
---

## Population Scale Analysis

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. 

```{r}
expr<-read.table
"rs8067378_ENSG00000172057.6.txt"
header = TRUE
```

```{r}
url <- "https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt"
expr <- read.table(url, header = TRUE)

head(expr)
```

```{r}
# Sample size per genotype
table(expr$geno)
```

```{r}
# Median expression per genotype
tapply(expr$exp, expr$geno, median)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
boxplot(exp ~ geno, data = expr,
        xlab = "Genotype",
        ylab = "Expression",
        col = c("#E74C3C", "#27AE60", "#2980B9"),
        notch = TRUE,
        main = "ORMDL3 Expression by Genotype")
```
Yes, the SNP appears to affect ORMDL3 expression. Individuals with the A/A genotype show higher expression levels than those with the G/G genotype.


