---
title: "Class 5: Data Viz with ggplot2"
author: "Hailey Phan (PID: A19113445)"
format: pdf
toc: true
---

## Background

There are many graphics systems in R for making plots and figures. These include so-called *"base R" graphics* like the `plot()` function and add on packages like **ggplot2**. 

Lets compare how we make a simple figure with these two systems:

We can use the in-built `cars` dataset:

```{r}
head(cars)
```

```{r}
plot(cars)
```

Before I can use ggplot2 I need to install it on my computer. To do this we can use the function `install.packages("ggplot2")`

> **N.B.** We never run `install.packages()` in our quarto doc (we run it only once in out R console) as. it would re-install the packages every time we render out quarto report

Once installed we need to load up the package into our R brain:

```{r}
library(ggplot2)
```


The main function in the **ggplot2** package is called `ggplot()`

```{r}
ggplot(cars)
```

Every ggplot has at least 3 layers:

- the **data** (a data.fram of the stuff we want to plot)
- the **aes**thetics (how the data maps to the plot)
- the **geom**etry layer (how you want the plot drawm, e.g. points, lines, etc.)

```{r}
ggplot(cars) + 
  aes(x = speed, y = dist) +
  geom_point()
```

## Add some custom features

Let's add a trend line that shows the relationship between speed and distance.

```{r}
ggplot(cars) + 
  aes(x = speed, y = dist) +
  geom_point() + 
  geom_line()
```

```{r}
ggplot(cars) + 
  aes(x = speed, y = dist) +
  geom_point() + 
  geom_smooth() +
  theme_bw() +
  labs(title = "Stoping Distance of Old Cars", 
       x = "Speed (MPH)",
       y = "Distance (ft)")
```

Q. Can you make the `geom_smooth()` function produce a linear straight line fit to the data and turn off the "gray" error region.

```{r}
ggplot(cars) + 
  aes(x = speed, y = dist) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() +
  labs(title = "Stoping Distance of Old Cars", 
       x = "Speed (MPH)",
       y = "Distance (ft)")
```

## Gene expression figure

Import the data to plot

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
sum(genes$State == "up")
```

A useful new function in this context is the `table()` function:

```{r}
table(genes$State)
```

My first plot attempt

```{r}
ggplot(genes) +
  aes(x = Condition1, Condition2, col = State) +
  geom_point()
```

Seond plot attempt

```{r}
p <- ggplot(genes) +
  aes(x = Condition1, Condition2, col = State) +
  geom_point()

p + scale_colour_manual( values=c("blue","gray","red") ) +
  labs(title = "Gene Expresion Changes Upon Drug Treatment",
         x = "Control (No Drug) ",
         y = "Drug Treatment")
```


## Going Further

Here we read the famous gapminder dataset

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder)
```


> Q. How many entries (i.e. rows) are in this dataset?

```{r}
nrow(gapminder)
```

> Q. How many different "country" entries are in this datset?

```{r}
length(table(gapminder$country))
```

```{r}
length(unique(gapminder$country))
```

Lets make out first plot of the entire dataset:

Plot of "gdpPercap" vs "lifeExp" colored by "continent"

```{r}
p <- ggplot(gapminder) +
  aes(x = gdpPercap, lifeExp, col = continent) +
  geom_point(alpha = 0.3)
```

I can add more layers to `p`

```{r}
p + 
  facet_wrap(~continent)
```

Make a plot for 1977 and 2007 only (not all the years in the dataset).

First use the **dplyr** package and the `fliter()` function from that package to extract the rows from the year 2007.

```{r}
library(dplyr)
```


```{r}
g07 <- filter(gapminder, year == 2007)

ggplot(g07) + 
  aes(x = gdpPercap, y = lifeExp, color = continent, size = pop) +
  geom_point()
```

```{r}
g97 <- filter(gapminder, year == 1997)

ggplot(g07) + 
  aes(x = gdpPercap, y = lifeExp, color = continent, size = pop) +
  geom_point()
```

```{r}
g07 <- filter(gapminder, year == 2007)
g77 <- filter (gapminder, year == 1977)
g <- filter(gapminder, year == 2007 | year == 1977)
```

```{r}
ggplot(g) +
  aes(x = gdpPercap, y = lifeExp, color = continent, size = pop) +
    geom_point() +
  facet_wrap(~year)
```


> Q. Make a histogram of lifeExp colored by continent (use `fill = continent` or `col = continent`)

```{r}
ggplot(gapminder) +
  aes(x = lifeExp, fill = continent) +
  geom_histogram() 
```

> Q. Make a histogram of lifeExp facted by continent

```{r}
ggplot(gapminder) +
  aes(lifeExp, fill = continent) +
  geom_histogram() +
  facet_wrap (~continent)
```

