---
title: "Class 13: Transcriptomics and the analysis of RNA-Seq data"
author: "Hailey Phan (PID: A19113445)"
toc: true
format: pdf
---

## Bioconductor setup

```{r}
library(BiocManager)
library(DESeq2)
```

## Import countData and colData

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <- read.csv("airway_metadata.csv")

head(counts)
head(metadata)
```
> Q1. How many genes are in this dataset? 

```{r}
nrow(counts)
```

> Q2. How many ‘control’ cell lines do we have? 

```{r}
sum(metadata$dex == "control")
```

## Toy differential gene expression

```{r}
library(dplyr)

control <- metadata %>% filter(dex == "control")
control.counts <- counts %>% select(control$id)
control.mean <- rowSums(control.counts) / nrow(control)
```


> Q3. How would you make the above code in either approach more robust? Is there a function that could help here? 

Instead of hardcoding `/4`, divide by `nrow(control)` or `nrow(treated)`. This way, if the number of samples changes, the code still works correctly without manual editing.

> Q4. Follow the same procedure for the `treated` samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called `treated.mean`)

```{r}
treated <- metadata %>% filter(dex == "treated")
treated.counts <- counts %>% select(treated$id)
treated.mean <- rowSums(treated.counts) / nrow(treated)
```

```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

> Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.

```{r}
plot(meancounts$control.mean, meancounts$treated.mean,
     xlab="Control Mean Counts",
     ylab="Treated Mean Counts")
```

> Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. What `geom_?()` function would you use for this plot? 

```{r}
library(ggplot2)
ggplot(meancounts, aes(x=control.mean, y=treated.mean)) +
  geom_point(alpha=0.3)
```
The geom function is geom_point().

> Q6. Try plotting both axes on a log scale. What is the argument to `plot()` that allows you to do this? 

```{r}
# Base R
plot(meancounts$control.mean, meancounts$treated.mean,
     log="xy",
     xlab="log Control Mean Counts",
     ylab="log Treated Mean Counts")

# ggplot version
ggplot(meancounts, aes(x=control.mean, y=treated.mean)) +
  geom_point(alpha=0.3) +
  scale_x_continuous(trans="log2") +
  scale_y_continuous(trans="log2")
```
The argument to plot() that enables log scaling on both axes is log="xy".


```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

> Q7. What is the purpose of the arr.ind argument in the `which()` function call above? Why would we then take the first column of the output and need to call the `unique()` function?

Setting `arr.ind=TRUE` makes `which()` return the result as a matrix of row and column indices (instead of just a flat index number), so you can identify exactly which row (gene) and column (sample group) contains a zero. We then take the first column to get just the row indices, and call `unique()` because a gene might have zeros in both columns. Without `unique()` we'd try to remove the same row twice, causing an error.

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```


> Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level? 

```{r}
sum(up.ind)
```

> Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level? 

```{r}
sum(down.ind)
```

> Q10. Do you trust these results? Why or why not?

No, this approach doesn't account for statistical significance. We're only looking at fold change in the means, with no consideration of variance, sample-to-sample variability, or sequencing depth differences. A gene could show a large fold change simply due to noise, especially at low counts. 

## Setting up for DESeq

```{r}
library(DESeq2)
citation("DESeq2")

dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

## Principal Component Analysis (PCA)

```{r}
vsd <- vst(dds, blind = FALSE)
plotPCA(vsd, intgroup = c("dex"))
```

```{r}
pcaData <- plotPCA(vsd, intgroup=c("dex"), returnData=TRUE)
head(pcaData)
```

```{r}
# Calculate percent variance per PC for the plot axis labels
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

```{r}
ggplot(pcaData) +
  aes(x = PC1, y = PC2, color = dex) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  theme_bw()
```

## DESeq analysis

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
res
```

```{r}
summary(res)
```

## Adding annotation data

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)
```

> Q11. Run the `mapIds()` function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called `res$entrez`, `res$uniprot` and `res$genename`.

```{r}
res$symbol  <- mapIds(org.Hs.eg.db, keys=row.names(res),
                      keytype="ENSEMBL", column="SYMBOL",    multiVals="first")
res$entrez  <- mapIds(org.Hs.eg.db, keys=row.names(res),
                      keytype="ENSEMBL", column="ENTREZID",  multiVals="first")
res$uniprot <- mapIds(org.Hs.eg.db, keys=row.names(res),
                      keytype="ENSEMBL", column="UNIPROT",   multiVals="first")
res$genename <- mapIds(org.Hs.eg.db, keys=row.names(res),
                       keytype="ENSEMBL", column="GENENAME", multiVals="first")

ord <- order(res$padj)
head(res[ord,])
write.csv(res[ord,], "deseq_results.csv")
```
## Data Visualization

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)
```

```{r}
# Setup our custom point color vector 
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

# Volcano plot with custom colors 
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

# Cut-off lines
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

```{r}
library(EnhancedVolcano)

x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```









