---
title: "Class 7: Machine Learning 1"
author: "Hailey Phan (PID: A19113445)"
format: pdf
toc: TRUE
---

## Background

Today we will explore some core machine learning methods that are very popular in bioinformatics. These include **clustering** and **dimensionallity reduction**.


## K-means clustering

The main function in "base" R for K-means clustering is called `kmeans()`.

Before we go too deep, let's make up some "simple" data that we can cluster and know if we are getting a good answer or not. To help us do this, we can use the `rnorm()` function:

```{r}
hist(rnorm(10000, mean = 3))
```

```{r}
rnorm(30, -3)
rnorm(30, +3)
```

```{r}
x <- c(rnorm(30,-3), rnorm(30, +3))
x
```

```{r}
z <- cbind(x = x, y = rev(x))
plot(z)
```

Now we can run `kmeans()` on this input `z` and see what the results look like.

```{r}
km <- kmeans(z, centers = 2)
km
```

```{r}
attributes(km)
```

> Q. How many points are in each cluster?

```{r}
km$size
```

> Q. What "component" of your result object details cluster assignment/membership?

```{r}
km$cluster
```


> Q. What "component" of your result object details cluster center?

```{r}
km$centers
```

> Q. Plot `z` colored by the kmeans cluster assignment and add cluster centers as blue points.

```{r}
plot(z, col = c("red", "blue"))
```

```{r}
plot(z, col = km$cluster)
points(km$centers, col = "blue", pch = 15)
```

> Q. Run a k-means clustering and plot the results asking for 4 clusters.

```{r}
km4 <- kmeans(z, centers = 4)

plot(z, col = km4$cluster)
points(km4$centers, col = "blue", pch = 15)
```

> **N.B.** You need to tell K-means the number of clusters (i.e. set `centers = 2`)!!

One approach is to try different values for `centers` and then pick the best...

```{r}
ans <- NULL
for(i in 1:10) {
  km <- kmeans(z, centers = i)
  ans <- c(ans, km$tot.withinss)
}

plot(ans, typ = "o", 
     xlab = "Number of clusters",
     ylab = "Total Sum of Squares")
```

## Hierarchical Clustering

The main function in "base" R for Heirarchical Clustering is called `hclust()`

This function does not take your "raw" data for clustering. You must first build a "distance matrix" from your data and pass this as input to `hclust()`

```{r}
d <- dist(z)
hc <- hclust(d)

hc
```

There is a bespoke `plot()` method for `hclust()` result objects.

```{r}
plot(hc)
```


Once we have out `hclust()` object (our "tree" of "cluster dendrogram") we can *"cut"* the tree to reveal the clustering pattern.

```{r}
cutree(hc, h = 8)
```

```{r}
cutree(hc, k = 4)
```

> Q. Make a plot of `z` with your `hclust()` results (i.e. colored by cluster membership).

```{r}

grps <- cutree(hc, k = 2)
plot(z, col = grps)
```


## Principal Component Analysis (PCA)

PCA is a dimensionallity reduction method that is popular for revealing patterns in complex datasets.



### Analysis of UK Food Data

Let's look at some data on the eating habits of folks from the UK to see if there are patterns that have some regions being distinct from others.

## Data Import

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```


> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```

```{r}
## Preview the first 6 rows
head(x)
```

```{r}
# Note how the minus indexing works
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```

```{r}
dim(x)
```

```{r}
x <- read.csv(url, row.names=1)
head(x)
```


> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

```{r}
read.csv(url, row.names = 1)
```

## Exploratory analysis

Make some plots to help make sense of obvious trends...

```{r}
# Using base R
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```


> Q3. Changing what optional argument in the above barplot() function results in the following plot?

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

## Tidy the data

Fix anything that 

```{r}
library(tidyr)
```


```{r}
# Convert data to long format for ggplot with `pivot_longer()`
x_long <- x |> 
          tibble::rownames_to_column("Food") |> 
          pivot_longer(cols = -Food, 
                       names_to = "Country", 
                       values_to = "Consumption")

dim(x_long)
```
```{r}
head(x_long)
```

```{r}
# Create grouped bar plot
library(ggplot2)

ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "dodge") +
  theme_bw()
```

> Q4. Changing what optional argument in the above ggplot() code results in a stacked barplot figure?

```{r}
library(ggplot2)

ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "stack") +
  theme_bw()
```

> Q5 We can use the `pairs()` function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```
> Points on the diagonal mean the variable is being plotted against itself, so they show a perfect correlation.


```{r}
library(pheatmap)

pheatmap( as.matrix(x) )
```

> Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?
> England, Wales, and Scotland cluster together. This suggests they have similar food consumption patterns. Northern Ireland is separated, meaning its diet differs. The main difference is higher consumption of items like fresh potatoes and soft drinks in Northern Ireland.


> **Key-point**: Even relatively small datasets can prove challenging to interpret.

## PCA to the rescue

The main function in "base" R for PCA is called `prcomp()`. This function wants the "observations" to be rows and the "variables" to be columns.

So here we need to take the transpose of our `x` input object.

```{r}
# Use the prcomp() PCA function 
pca <- prcomp(t(x))
summary(pca)
```

> Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
# Create a data frame for plotting
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

# Plot PC1 vs PC2 with ggplot
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

> Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

```{r}
ggplot(df) +
  aes(x = PC1, y = PC2, label = Country, color = Country) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  scale_color_manual(values = c(
    "England" = "orange",
    "Wales" = "red",
    "Scotland" = "blue",
    "N.Ireland" = "green"
  )) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
```

```{r}
## or the second row here...
z <- summary(pca)
z$importance
```

```{r}
# Create scree plot with ggplot
variance_df <- data.frame(
  PC = factor(paste0("PC", 1:length(v)), levels = paste0("PC", 1:length(v))),
  Variance = v
)

ggplot(variance_df) +
  aes(x = PC, y = Variance) +
  geom_col(fill = "steelblue") +
  xlab("Principal Component") +
  ylab("Percent Variation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0))
```

```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```

> Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

```{r}
ggplot(pca$rotation) +
  aes(x = PC2, y = reorder(rownames(pca$rotation), PC2)) +
  geom_col(fill = "steelblue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw()
```
> PC2 mainly reflects differences in types of vegetable consumption, especially between fresh and processed foods. It separates countries based on whether they consume more fresh vegetables versus processed vegetables.
