---
title: "Lab 10: Structural Bioinformatics"
author: "Hailey Phan (PID: A19113445)"
format: pdf
toc: true
prefer-html: true
---

## PDB statistics

```{r}
pdb_stats <- read.csv("pdb_stats.csv")
head(pdb_stats)
```

> Q1: What percentage of structures in the PDB are solved by X-Ray and Electron Microscopy.

```{r}
total <- 253482
xray <- 203939
em <- 34005

((xray + em) / total) * 100
```
About 93.87% of PDB structures are solved by X-ray or Electron Microscopy.

> Q2: What proportion of structures in the PDB are protein?

```{r}
total <- 253482

protein_only <- 180891 + 23208 + 12816 + 350 + 229 + 84
protein_oligo <- 10495 + 3745 + 34 + 8 + 11 + 1
protein_na <- 9206 + 6773 + 287 + 26 + 8 + 0

protein_total <- protein_only + protein_oligo + protein_na

protein_total / total
(protein_total / total) * 100
```

The proportion of PDB structures that are protein is about 0.979, or 97.91%

> Q3: Type HIV in the PDB website search box on the home page and determine how many HIV-1 protease structures are in the current PDB?

There are currently 1,227 HIV-1 protease structures in the PDB.


## The important role of water

> Q4: Water molecules normally have 3 atoms. Why do we see just one atom per water molecule in this structure?

Water molecules normally have 3 atoms, but in X-ray crystal structures the hydrogen atoms are usually not visible because they scatter X-rays very weakly. Because of this, each water molecule is usually shown as only the oxygen atom.

> Q5: There is a critical “conserved” water molecule in the binding site. Can you identify this water molecule? What residue number does this water molecule have?

The conserved water molecule in the binding site is water molecule HOH 308. This water molecule is located between the ligand and the flap residues near Ile-50 in the active site and helps stabilize ligand binding through hydrogen bonding interactions.

> Q6

![HIV-1 protease (1HSG) showing two chains, indinavir ligand, 
ASP25 residues and conserved water HOH308](1HSG.png)


## Introduction to Bio3D in R

```{r}
library(bio3d)
```

```{r}
pdb <- read.pdb("1hsg")
pdb
```

> Q7: How many amino acid residues are there in this pdb object? 

198 residues

> Q8: Name one of the two non-protein residues? 

HOH

> Q9: How many protein chains are in this structure? 

2 chains

```{r}
attributes(pdb)
```

```{r}
head(pdb$atom)
```

```{r}
library(bio3dview)
library(NGLVieweR)

#view.pdb(pdb) |>
  #setSpin()
```

```{r}
# Select the important ASP 25 residue
sele <- atom.select(pdb, resno=25)

# and highlight them in spacefill representation
#view.pdb(pdb, cols=c("navy","teal"), 
         #highlight = sele,
         #highlight.style = "spacefill") |>
  #setRock()
```

```{r}
adk <- read.pdb("6s36")
adk
```

```{r}
# Perform flexiblity prediction
m <- nma(adk)
plot(m)
```

```{r}
#mktrj(m, file="adk_m7.pdb")
#view.nma(m, pdb=adk)
```
## Comparative structure analysis of Adenylate Kinase

> Q10. Which of the packages above is found only on BioConductor and not CRAN? 

msa.

> Q11. Which of the above packages is not found on BioConductor or CRAN?: 

bio3dview.

> Q12. True or False? Functions from the pak package can be used to install packages from GitHub and BitBucket? 

True.

```{r}
library(bio3d)

# Get the ADK sequence
aa <- get.seq("1ake_A")
aa
```

> Q13. How many amino acids are in this sequence, i.e. how long is this sequence? 

214 amino acids

```{r}
hits <- NULL
hits$pdb.id <- c('1AKE_A','6S36_A','6RZE_A','3HPR_A','1E4V_A','5EJE_A','1E4Y_A','3X2S_A','6HAP_A','6HAM_A','4K46_A','3GMT_A','4PZL_A')
```

```{r}
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)

```

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```
## Annotate collected PDB structures

```{r}
# Vector containing PDB database codes
ids <- basename.pdb(pdbs$id)

anno <- pdb.annotate(ids)
unique(anno$source)
```

```{r}
# Perform PCA
pc.xray <- pca(pdbs)
plot(pc.xray)
```

```{r}
# Calculate RMSD
rd <- rmsd(pdbs)

# Structure-based clustering
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)

plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```

```{r}
# Visualize first principal component
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```

```{r}
#Plotting results with ggplot2
library(ggplot2)
library(ggrepel)

df <- data.frame(PC1=pc.xray$z[,1], 
                 PC2=pc.xray$z[,2], 
                 col=as.factor(grps.rd),
                 ids=ids)

p <- ggplot(df) + 
  aes(PC1, PC2, col=col, label=ids) +
  geom_point(size=2) +
  geom_text_repel(max.overlaps = 20) +
  theme(legend.position = "none")
p
```





