---
title: "Class 14: Pathway Analysis from RNA-Seq Results"
author: "Hailey Phan (PID: A19113445)"
format: pdf
toc: true
---

## Section 1. Differential Expression Analysis

```{r}
library(DESeq2)

# Load data
colData = read.csv("GSE37704_metadata.csv", row.names=1)
countData = read.csv("GSE37704_featurecounts.csv", row.names=1)

head(colData)
head(countData)
```
> Q. Complete the code below to remove the troublesome first column from countData

```{r}
countData <- as.matrix(countData[, -1])
head(countData)
```

> Q. Complete the code below to filter countData to exclude genes (i.e. rows) where we have 0 read count across all samples (i.e. columns).

```{r}
countData = countData[rowSums(countData) > 0, ]
head(countData)
```


# Running DESeq2

```{r}
dds = DESeqDataSetFromMatrix(countData = countData,
                             colData = colData,
                             design = ~condition)
dds = DESeq(dds)
dds
```

```{r}
res = results(dds)
```

> Q. Call the summary() function on your results to get a sense of how many genes are up or down-regulated at the default 0.1 p-value cutoff.

```{r}
summary(res)
```

# Volcano Plot

```{r}
library(ggplot2)

# Convert to data frame for ggplot
res_df <- as.data.frame(res)

ggplot(res_df) +
  aes(x = log2FoldChange,
      y = -log10(padj)) +
  geom_point()
```

> Q. Improve this plot by completing the below code, which adds color, axis labels and cutoff lines:


```{r}
# Color vector
mycols <- rep("gray", nrow(res_df))

# Blue for abs(log2FC) > 2
mycols[abs(res_df$log2FoldChange) > 2] <- "blue"

# Gray back those with padj > 0.01
mycols[res_df$padj > 0.01] <- "gray"

ggplot(res_df) +
  aes(x = log2FoldChange,
      y = -log10(padj)) +
  geom_point(color = mycols) +
  xlab("Log2(FoldChange)") +
  ylab("-Log(P-value)") +
  geom_vline(xintercept = c(-2, 2), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed")
```

# Adding gene annotation

> Q. Use the mapIDs() function multiple times to add SYMBOL, ENTREZID and GENENAME annotation to our results by completing the code below.

```{r}
library(AnnotationDbi)
library(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                    keys = row.names(res),
                    keytype = "ENSEMBL",
                    column = "SYMBOL",
                    multiVals = "first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys = row.names(res),
                    keytype = "ENSEMBL",
                    column = "ENTREZID",
                    multiVals = "first")

res$name = mapIds(org.Hs.eg.db,
                  keys = row.names(res),
                  keytype = "ENSEMBL",
                  column = "GENENAME",
                  multiVals = "first")

head(res, 10)
```

>Q. Finally for this section let's reorder these results by adjusted p-value and save them to a CSV file in your current project directory.

```{r}
res = res[order(res$pvalue), ]
write.csv(res, file = "deseq_results.csv")
```

## Section 2. Pathway Analysis

```{r}
library(pathview)
library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

# Keep only signaling and metabolic pathways
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Named fold change vector
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)

# Run GAGE
keggres = gage(foldchanges, gsets = kegg.sets.hs)
attributes(keggres)

# Top down-regulated pathways
head(keggres$less)
```

```{r}
# Plot top down-regulated pathway (Cell cycle)
pathview(gene.data = foldchanges, pathway.id = "hsa04110")

# PDF version
pathview(gene.data = foldchanges, pathway.id = "hsa04110", kegg.native = FALSE)
```

```{r}
# Plot top down-regulated pathway (Cell cycle)
pathview(gene.data = foldchanges, pathway.id = "hsa04110")

# PDF version
pathview(gene.data = foldchanges, pathway.id = "hsa04110", kegg.native = FALSE)
```

> Q. Can you do the same procedure as above to plot the pathview figures for the top 5 down-regulated pathways?

```{r}
# Get top 5 down-regulated pathway IDs
keggrespathways_down <- rownames(keggres$less)[1:5]
keggresids_down <- substr(keggrespathways_down, start = 1, stop = 8)
keggresids_down

# Plot all 5
pathview(gene.data = foldchanges, pathway.id = keggresids_down, species = "hsa")
```

## Section 3. Gene Ontology (GO)

```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets = gobpsets)

lapply(gobpres, head)
```

## Section 4. Reactome Analysis

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))

write.table(sig_genes, file = "significant_genes.txt",
            row.names = FALSE, col.names = FALSE, quote = FALSE)
```

> Q: What pathway has the most significant “Entities p-value”? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?

The most significant Reactome pathway is typically Cell Cycle (specifically mitotic processes), which aligns well with the KEGG results showing Cell cycle and DNA replication as top down-regulated pathways. Differences between the two methods can arise because: Reactome and KEGG use different pathway definitions and curation approaches; Reactome tends to have more granular/nested pathways; and the two databases have different gene coverage and annotation depths.






